---
title: "lmds: Landmark Multi-Dimensional Scaling"
author: rcannood
date: 2019-11-13T20:53:00+00:00
categories: ["Algorithms"]
tags: ["Algorithms"]
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(collapse = TRUE)
```

Multi-dimensional scaling (MDS) is a useful algorithm for visualising high-dimensional data in a
low-dimensional space. A major benefit of MDS in comparison to linear transformation methods
such as principal component analysis (PCA) is that it allows using a custom defined distance metric, 
not just euclidean. 

```{r pcavsmds}
library(tidyverse)
x <- as.matrix(dyno::fibroblast_reprogramming_treutlein$expression)
group <- dyno::fibroblast_reprogramming_treutlein$grouping
x <- sweep(x, MARGIN = 1, runif(nrow(x), 1, 100), "*")

dimred_pca <- prcomp(x)$x
dimred_mds <- cmdscale(.5 - cor(t(x), method = "spearman") / 2)

qplot(dimred_pca[,1], dimred_pca[,2], colour = group)
qplot(dimred_mds[,1], dimred_mds[,2], colour = group)
```
...WIP...

A while back, lmds v0.1.0 was published on [CRAN](https://cran.r-project.org/package=lmds).

Landmark Multi-Dimensional Scaling (lmds) is an extension of classical Torgerson MDS.
LMDS aims to produce the same dimensionality reduction as `cmdscale()`, but scale
linearly with respect to the number of samples.

```{r compare}
library(lmds)

x <- as.matrix(iris[,1:4])
dimred <- lmds(x, ndim = 2)
qplot(dimred[,1], dimred[,2]) + labs(title = "lmds()") + theme_classic()

dimred <- cmdscale(dist(x))
qplot(dimred[,1], dimred[,2]) + labs(title = "cmdscale()") + theme_classic() 
```


## Execution time
The execution time of `lmds()` scales linearly with respect to the dataset size.
```{r timings, echo=FALSE}
largex <- Matrix::rsparsematrix(nrow = 100000, ncol = 10000, density = .01)

log <- list()
num_samples <- 100
time <- 0
while (time < 10) {
  subx <- largex[seq_len(num_samples),]
  time <- system.time({
    dimred <- lmds(subx)
  })[["user.self"]]
  log[[length(log) + 1]] <- tibble(method = "lmds()", num_samples, time)
  num_samples <- num_samples * 1.5
}

num_samples <- 100
time <- 0
while (time < 10) {
  subx <- largex[seq_len(num_samples),]
  time <- system.time({
    dist <- dist(subx)
    dimred <- cmdscale(dist)
  })[["user.self"]]
  log[[length(log) + 1]] <- tibble(method = "cmdscale()", num_samples, time)
  num_samples <- num_samples * 1.5
}

logdf <- bind_rows(log)

ggplot(logdf, aes(num_samples, time, colour = method)) + 
  geom_point() +
  geom_line() +
  theme_classic()
```
